---
title: "work_stanley"
output: html_document
---

## Libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(ggplot2)
library(naniar)
library(ggmap)
library(shiny)
library(shinydashboard)
library(UKFE)
library(bslib)
```

## Data
We will use data from [William B.V. Langdon, Richard Fox, Owen T. Lewis, Host plant use is driven by microclimate not nutritional quality in a grassland butterfly](https://doi.org/10.1101/2025.11.15.688506). The original research article and data sets are included in the data folder in this repository.
```{r}
egg_laying <- read_csv("data/egg_laying_data.csv") %>% 
  clean_names()
plant_chemistry <- read_csv("data/plant_chemistry_data.csv") %>%
  clean_names()
```

## Exploratory Data Analysis
**Looking Into Egg Laying Data Set**

```{r}
names(egg_laying)
```

```{r}
summary(egg_laying) # look at overall distribution of data, check for abnormal values
```

```{r}
egg_laying %>% 
  miss_var_summary() #check for NA's
```

```{r}
egg_laying %>% 
  gg_miss_var()+ #visualizing the NA's
  labs(title="NA's in Egg Laying Data Set")
```

Through the NA summary and plot, we can see the majority of NA's come from environment_date with most of the other NA's from leaf2 and leaf3 measurements. It's possible that they left the environmental measurement dates out on accident as variables besides ambient and soil temperature were measured despite the date being NA. For the leaf measurements, it could be that researchers thought that 1 or 2 leaf measurements of a plant seemed representative of the rest of the plant's leaves.

**Looking Into Plant Chemistry Data Set**

```{r}
names(plant_chemistry)
```

```{r}
summary(plant_chemistry)
```

```{r}
plant_chemistry %>% 
  miss_var_summary() #check for NA's
```

The data set, plant_chemistry, is clean and tidy without any NA's or suspicious values.

**Joining the Data Sets**

Since the egg_laying and plant_chemistry data sets have the same plant id, we can join the two data sets together using a join function in case we want to use variables of a plant id from both data sets. It's important to know that plant_chemistry also shares the same leaf length, width, area, plant area, and season, so we can select the unique variables of plant_chemistry.

```{r}
plant_joined <- left_join(egg_laying, plant_chemistry %>% select(plant_id, n, c, cn_percent, n_molar, c_molar, cn_molar, quality_pc1, quality_pc2), by="plant_id")
```

**Distribution of Important Variables**

```{r}
plant_joined %>% 
  select(season) %>% 
  count(season, sort=T) 
```

```{r}
plant_joined %>% 
  select(season) %>% 
  ggplot(aes(x=season))+ 
  geom_bar(aes(fill=season), color="black")+
  labs(title="Number of Observations in Each Season", x="Season", y=NULL, fill="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(season, predicted_n) %>% 
  ggplot(aes(x=season, y=predicted_n))+
  geom_boxplot(aes(fill=season))+
  labs(title="Range of Predicted Nitrogen in Each Season", x="Season", y="Predicted Plant Nitrogen (micromole/gram)", fill="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(season, temperature_soil) %>% 
  ggplot(aes(x=season, y=temperature_soil))+
  geom_boxplot(aes(fill=season))+
  labs(title="Range of Soil Temperature in Each Season", x="Season", y="Soil Temperature (degrees Celsius)", fill="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(season, temperature_ambient) %>% 
  ggplot(aes(x=season, y=temperature_ambient))+
  geom_boxplot(aes(fill=season))+
  labs(title="Range of Ambient Temperature in Each Season", x="Season", y="Ambient Temperature (degrees Celsius)", fill="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(season, average_leaf_size) %>% 
  ggplot(aes(x=season, y=average_leaf_size))+
  geom_boxplot(aes(fill=season))+
  labs(title="Range of Average Leaf Size in Each Season", x="Season", y="Average Leaf Size (mm^2)", fill="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(season, plant_area) %>% 
  ggplot(aes(x=season, y=plant_area))+
  geom_boxplot(aes(fill=season))+
  labs(title="Range of Plant Area in Each Season", x="Season", y="Plant Area (cm^2)", fill="Season")+
  theme_minimal()
```

**Interpreting Paper's Principal Component Analysis**

Principal Component Analysis is a dimension-reducing strategy that converts highly correlated variables into non-correlated variables called "principal components". These principal components retain the original data set's variability. Our data set comes with two variables, quality_PC1 and quality_PC2, which are values of principal components that the researchers of the paper got to represent average leaf size and plant size. We can attempt to visualize the principal components' relationships to other variables.
```{r}
plant_joined %>% 
  select(bare_ground, quality_pc1, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=quality_pc1, y=bare_ground))+
  geom_point(aes(color=season))+
  geom_smooth(method="lm", se=T)+
  facet_wrap(~season, ncol=2, labeller=label_both)+
  labs(title="Principal Component 1 vs. %Bare Ground", x="Principal Component 1", y="%Bare Ground", color="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(bare_ground, quality_pc2, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=quality_pc2, y=bare_ground))+
  geom_point(aes(color=season))+
  geom_smooth(method="lm", se=T)+
  facet_wrap(~season, ncol=2, labeller=label_both)+
  labs(title="Principal Component 2 vs. %Bare Ground", x="Principal Component 2", y="%Bare Ground", color="Season")+
  theme_minimal()
```

From these plots, we can see that bare ground percentage is negatively correlated with both principal components relating to the average plant and leaf size.

```{r}
plant_joined %>% 
  select(veg, quality_pc1, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=quality_pc1, y=veg))+
  geom_point(aes(color=season))+
  geom_smooth(method="lm", se=T)+
  facet_wrap(~season, ncol=2, labeller=label_both)+
  labs(title="Principal Component 1 vs. %Veg", x="Principal Component 1", y="%Veg", color="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(veg, quality_pc2, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=quality_pc2, y=veg))+
  geom_point(aes(color=season))+
  geom_smooth(method="lm", se=T)+
  facet_wrap(~season, ncol=2, labeller=label_both)+
  labs(title="Principal Component 2 vs. %Veg", x="Principal Component 2", y="%Veg", color="Season")+
  theme_minimal()
```

From these plots, we see that vegetation percentage is positively correlated with both principal components.

```{r}
plant_joined %>% 
  select(litter, quality_pc1, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=quality_pc1, y=litter))+
  geom_point(aes(color=season))+
  geom_smooth(method="lm", se=T)+
  facet_wrap(~season, ncol=2, labeller=label_both)+
  labs(title="Principle Component 1 vs. %Leaf Litter", x="Principle Component 1", y="%Leaf Litter", color="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(litter, quality_pc2, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=quality_pc2, y=litter))+
  geom_point(aes(color=season))+
  geom_smooth(method="lm", se=T)+
  facet_wrap(~season, ncol=2, labeller=label_both)+
  labs(title="Principle Component 2 vs. %Leaf Litter", x="Principle Component 2", y="%Leaf Litter", color="Season")+
  theme_minimal()
```

Interestingly, we don't see the same consistent correlations with surrounding leaf litter percentage against both principle components. We see that in autumn, the litter percentage has no correlation between principle component 1 and 2 while in spring, the litter percentage is postive correlated with principle component 1 and negatively correlated with principle component 2.

**Creating Map for Data Set**

```{r}
register_stadiamaps("f9de653c-55a3-4f5b-9672-63291557e012", write = FALSE) # will probably take this down after repository check
coordtomap <- egg_laying %>% 
  select(gps, season) %>% 
  na.omit() %>% 
  separate(gps, into=c("SP", "grid"), sep=2) %>% # separate numerical part from SP 
  separate(grid, into=c("easting", "northing"), sep=5) %>% # divide number into easting and northing portions
  mutate(easting=as.numeric(easting)+400000) %>% # convert easting into latitude
  mutate(northing=as.numeric(northing)+200000) %>% #convert nothing into longitude
  rowwise() %>% # treat rows as separate groups in order to apply the ConvertGridRef to each row
  mutate(coords=list(ConvertGridRef(c(easting,northing)))) %>% 
  unnest(coords) # converts latitude and longitude from the list as separate variables
```

```{r}
summary(coordtomap) # find the minimum and maximum latitude and longitude
```

```{r}
lat <- c(51.74 , 51.75) # 
long <- c(-1.185 , -1.181)
bbox <- make_bbox(long, lat, f=0.25) # create bounding box
map1 <- get_stadiamap(bbox, maptype="stamen_terrain", zoom=15)
ggmap(map1)+ # creating the map itself
  geom_point(data=coordtomap, aes(longitude, latitude, color=season), size=0.75, alpha=0.8)+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(title="Distribution of Plants Around Shotover Country Park (United Kingdom)", x="Longitude", y="Latitude", color="Season")
```

## Factors Associated with Plant Nutritional Quality

Plant nutritional quality is dependent on the chemical composition of each plant, with butterflies usually benefiting from high nitrogen-containing plants so that they don't need to overeat to obtain sufficient nutrients. Plant nitrogen is also representative of its soil nitrogen and improves plant growth in the vicinity. Therefore, having high nitrogen would lead to taller plants which then would lead to a lower soil temperature because less sun is able to reach the bottom of plants.

We can check to see if each of these correlations are true. We should expect to see a positive association between plant nitrogen and percent of vegetation around a plant, a positive association between plant nitrogen and sward height, which is the height that approximately 80% of plants grew at, and a negative association between predicted plant nitrogen and soil temperature.

**Plot of Predicted Plant Nitrogen vs. %Vegetation**
```{r}
plant_joined %>% 
  select(predicted_n, veg, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=veg))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. %Vegetation", x="Predicted Plant Nitrogen (micromole/gram)", y="%Vegetation", color="Season")+
  theme_minimal()
```


**Plot of Predicted Plant Nitrogen vs. Sward Height**
```{r}
plant_joined %>%
  select(predicted_n, sward_height, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=sward_height))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. Sward Height", x="Predicted Plant Nitrogen (micromole/gram)", y="Sward Height (cm)", color="Season")+
  theme_minimal()
```

**Plot of Predicted Plant Nitrogen vs. Soil Temperature**
```{r}
plant_joined %>% 
  select(predicted_n, temperature_soil, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=temperature_soil))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. Soil Temperature", x="Predicted Plant Nitrogen (micromol/g)", y="Soil Temperature (°C)", color="Season")+
  theme_minimal()
```

**Discussion About Vegetation-Related Plant Nitrogen Plots**

From the plots, we see that the correlation of plant nitrogen and vegetation % or sward height is actually negative. The only correlation that matches our prediction is the negative correlation of plant nitrogen and soil temperature. The different seasons in each plot occupy a different area, with plants in spring usually having a larger vegetation area, taller plant height, and greater soil temperature at a significantly lower plant nitrogen than plants in autumn.

**Leaf and Plant Size as a Possible Indicator of Plant Nitrogen**

One of the important uses of nitrogen for plants is the creation of chlorophyll, a green pigment required for photosynthesis. Having a larger leaf area could mean a higher total amount of chlorophyll and a higher nitrogen content in a plant. A larger plant would also have a larger biomass and would require more nitrogen from the soil. We can test whether leaf and plant size are positive associated with plant nitrogen content.

```{r}
plant_joined %>% 
  select(predicted_n, total_leaf_area, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=total_leaf_area))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. Average Leaf Area", x="Predicted Plant Nitrogen (micromole/gram)", y="Average Leaf Area (mm^2)", color="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(predicted_n, plant_area, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=plant_area))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. Plant Area", x="Predicted Plant Nitrogen (micromole/gram)", y="Plant Area (cm^2)", color="Season")+
  theme_minimal()
```


**Discussion About Leaf and Plant Size**

From the plots, we can see a strong correlation between plant nitrogen content and leaf area or plant area. These results make leaf area and plant area a good predictor of plant nitrogen and, therefore, plant nutrition for butterflies.


## Egg Amount Predictor Shiny App
All of our analyses on our data set's variables have been bivariate so far. However, bivariate analyses don't account for other variables' influences on a dependent variable in its correlation. Using multiple regression analysis, we can use multiple predictor variables to predict a dependent variable and see the strength each variable has in influencing it.

**Creating the Regression Formula for App**
```{r}
egg_calculator <- lm(formula=n_eggs ~ predicted_n + bare_ground + veg + sward_height + plant_area + average_leaf_size + temperature_ambient + temperature_soil, data=plant_joined) # create a multiple regression formula
egg_calculator_residuals <- egg_calculator$residuals
```

**Create QQ Plot**

A QQ Plot can be used in conjunction with a multiple regression analysis to see whether its residuals are normal or not. In the case of this analysis, our data is approximately normal with some skewedness to the right as the plot skews upward.
```{r}
qqnorm(egg_calculator_residuals)
qqline(egg_calculator_residuals)
```

**Print Multiple Regression Model Summary**

We can see the estimated effect each variable of our regression model has on our dependent variable, number of eggs. We can see that in the summary, bare ground, sward height, and soil temperature are strong estimators of the number of eggs as its p values are significantly low, as indicated by the asterisks.
```{r}
summary(egg_calculator)
```

**Code of Actual Shiny Code**
```{r}

ui <- page_sidebar(
  title = "Egg Calculator Regression",
  sidebar = sidebar(
    
    # Select predictor variables
    checkboxGroupInput(
      "predictor_vars",
      "Select Predictor Variables:",
      choices = c("predicted_n",
                  "bare_ground",
                  "veg",
                  "sward_height",
                  "plant_area",
                  "average_leaf_size",
                  "temperature_ambient",
                  "temperature_soil"
      ),
      selected = c("predicted_n")
    ),
    
    # Action button to run regression
    actionButton("run_regression", "Run Regression", class = "btn-primary mt-3")
  ),
  
  # Main panel with plots and results
  navset_card_tab(
    nav_panel(
      "Scatter Plots",
      card(
        card_header("Egg Count vs Predictor Variables"),
        plotOutput("scatter_plots", height = "600px")
      )
    ),
    nav_panel(
      "Regression Summary",
      card(
        card_header("Model Summary"),
        verbatimTextOutput("regression_summary")
      ),
      card(
        card_header("Model Coefficients"),
        tableOutput("coefficients_table")
      )
    ),
    nav_panel(
      "Diagnostic Plots",
      card(
        card_header("Residual Diagnostics"),
        plotOutput("diagnostic_plots", height = "600px")
      )
    )
  )
)

server <- function(input, output, session) {
  
  # Fit regression model when button is clicked
  model <- eventReactive(input$run_regression, {
    req(input$predictor_vars)
    
    # Create formula with egg count as dependent variable
    formula_str <- paste("n_eggs ~", paste(input$predictor_vars, collapse = " + "))
    formula_obj <- as.formula(formula_str)
    
    # Fit linear model
    lm(formula_obj, data = plant_joined)
  })
  
  # Scatter plots
  output$scatter_plots <- renderPlot({
    req(input$predictor_vars)
    
    n_predictors <- length(input$predictor_vars)
    
    # Calculate grid dimensions
    n_cols <- ceiling(sqrt(n_predictors))
    n_rows <- ceiling(n_predictors / n_cols)
    
    par(mfrow = c(n_rows, n_cols), mar = c(4, 4, 2, 1))
    
    for (predictor in input$predictor_vars) {
      plot(plant_joined[[predictor]], plant_joined$n_eggs,
           xlab = predictor,
           ylab = "Egg Count",
           main = paste("Egg Count vs", predictor),
           pch = 19,
           col = rgb(0.2, 0.4, 0.6, 0.6),
           cex = 1.2)
      
      # Add regression line for visualization
      abline(lm(plant_joined$n_eggs ~ plant_joined[[predictor]]), 
             col = "red", lwd = 2)
    }
  })
  
  # Regression summary
  output$regression_summary <- renderPrint({
    req(model())
    summary(model())
  })
  
  # Coefficients table
  output$coefficients_table <- renderTable({
    req(model())
    
    coef_summary <- summary(model())$coefficients
    coef_df <- as.data.frame(coef_summary)
    coef_df$Variable <- rownames(coef_df)
    coef_df <- coef_df[, c("Variable", "Estimate", "Std. Error", "t value", "Pr(>|t|)")]
    rownames(coef_df) <- NULL
    
    coef_df
  }, digits = 4)
  
  # Diagnostic plots
  output$diagnostic_plots <- renderPlot({
    req(model())
    
    par(mfrow = c(2, 2))
    plot(model())
  })
}

shinyApp(ui, server)
```


## Limitations Of Data and Correlations
The data collected and used for the correlations in our study is specifically from the Shotover Country Park in Oxfordshire, United Kingdom. The only the plant species used in this data was Rumex acetosa or sorrel, which limits the study to butterflies that specialize in consuming this plant, specifically Lycaena phlaeas. This may impact the validity of our study because the data sample isn't representative of plants and butterflies throughout the world and our study's findings may not generalize to them. Future studies on the trade-off between plant nutritional value and egg-laying preference could be done in other areas of the world, on other species of plants, and other species of butterflies.