---
title: "work_stanley"
output: html_document
---

## Libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(ggplot2)
library(naniar)
library(ggmap)
library(shiny)
library(shinydashboard)
library(UKFE)
library(bslib)
```

## Data
We will use data from [William B.V. Langdon, Richard Fox, Owen T. Lewis, Host plant use is driven by microclimate not nutritional quality in a grassland butterfly](https://doi.org/10.1101/2025.11.15.688506). The original research article and data sets are included in the data folder in this repository.
```{r}
egg_laying <- read_csv("data/egg_laying_data.csv") %>% 
  clean_names()
plant_chemistry <- read_csv("data/plant_chemistry_data.csv") %>%
  clean_names()
```

## Exploratory Data Analysis
**Looking Into Egg Laying Data Set**

```{r}
names(egg_laying)
```

```{r}
summary(egg_laying) # look at overall distribution of data, check for abnormal values
```

```{r}
egg_laying %>% 
  miss_var_summary() #check for NA's
```

```{r}
egg_laying %>% 
  gg_miss_var()+ #visualizing the NA's
  labs(title="NA's in Egg Laying Data Set")
```

**Looking Into Plant Chemistry Data Set**

```{r}
names(plant_chemistry)
```

```{r}
summary(plant_chemistry)
```

```{r}
plant_chemistry %>% 
  miss_var_summary() #check for NA's
```

The data set, plant_chemistry, is clean and tidy without any NA's or suspicious values.

**Joining the Data Sets**

Since the egg_laying and plant_chemistry data sets have the same plant id, we can join the two data sets together using a join function. It's important to know that plant_chemistry also shares the same leaf length, width, area, plant area, and season, so we can select the unique variables of plant_chemistry.

```{r}
plant_joined <- left_join(egg_laying, plant_chemistry %>% select(plant_id, n, c, cn_percent, n_molar, c_molar, cn_molar, quality_pc1, quality_pc2), by="plant_id")
```

**Distribution of Important Variables**

```{r}
plant_joined %>% 
  select(season) %>% 
  count(season, sort=T) #obtain counts of season numerically
```

```{r}
plant_joined %>% 
  select(season) %>% 
  ggplot(aes(x=season))+ # visually compare the counts of season in this data set
  geom_bar(aes(fill=season))+
  labs(title="Number of Observations in Each Season", x="Season", y=NULL, fill="Season")+
  theme_minimal()
```

There are more observations done during the spring season than autumn (a difference of 50 and a spring:autumn ratio of 1.37:1), but it's likely not significant enough to skew the data.

**

**Creating Map for Data Set**

```{r}
register_stadiamaps("f9de653c-55a3-4f5b-9672-63291557e012", write = FALSE)
coordtomap <- egg_laying %>% 
  select(gps, season) %>% 
  na.omit() %>% 
  separate(gps, into=c("SP", "grid"), sep=2) %>% # separate numerical part from SP 
  separate(grid, into=c("easting", "northing"), sep=5) %>% # divide number into easting and northing portions
  mutate(easting=as.numeric(easting)+400000) %>% # convert easting into latitude
  mutate(northing=as.numeric(northing)+200000) %>% #convert nothing into longitude
  rowwise() %>% # treat rows as separate groups in order to apply the ConvertGridRef to each row
  mutate(coords=list(ConvertGridRef(c(easting,northing)))) %>% 
  unnest(coords) # converts latitude and longitude from the list as separate variables
```

```{r}
summary(coordtomap) # find the minimum and maximum latitude and longitude
```

```{r}
lat <- c(51.74 , 51.75) # 
long <- c(-1.185 , -1.181)
bbox <- make_bbox(long, lat, f=0.25) # create bounding box
map1 <- get_stadiamap(bbox, maptype="stamen_terrain", zoom=15)
ggmap(map1)+
  geom_point(data=coordtomap, aes(longitude, latitude, color=season), size=0.75, alpha=0.8)+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(title="Distribution of Plants Around Shotover Country Park (United Kingdom)", x="Longitude", y="Latitude", color="Season")
```

```{r}
ggmap(map1)+
  geom_point(data=coordtomap, aes(longitude, latitude, color=season), size=0.75, alpha=0.8)+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(title="Distribution of Plants Around Shotover Country Park (United Kingdom)", x="Longitude", y="Latitude", color="Season")
```


## Factors Associated with Plant Nutritional Quality

Plant nutritional quality is dependent on the chemical composition of each plant, with butterflies usually benefiting from high nitrogen-containing plants so that they don't need to overeat to obtain sufficient nutrients. Plant nitrogen is also representative of its soil nitrogen and improves plant growth in the vicinity. Therefore, having high nitrogen would lead to taller plants which then would lead to a lower soil temperature because less sun is able to reach the bottom of plants.

We can check to see if each of these correlations are true. We should expect to see a positive association between plant nitrogen and percent of vegetation around a plant, a positive association between plant nitrogen and sward height, which is the height that approximately 80% of plants grew at, and a negative association between predicted plant nitrogen and soil temperature.

**Plot of Predicted Plant Nitrogen vs. %Vegetation**
```{r}
plant_joined %>% 
  select(predicted_n, veg, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=veg))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. %Vegetation", x="Predicted Plant Nitrogen (micromole/gram)", y="%Vegetation", color="Season")+
  theme_minimal()
```


**Plot of Predicted Plant Nitrogen vs. Sward Height**
```{r}
plant_joined %>%
  select(predicted_n, sward_height, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=sward_height))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. Sward Height", x="Predicted Plant Nitrogen (micromole/gram)", y="Sward Height (cm)", color="Season")+
  theme_minimal()
```

**Plot of Predicted Plant Nitrogen vs. Soil Temperature**
```{r}
plant_joined %>% 
  select(predicted_n, temperature_soil, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=temperature_soil))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. Soil Temperature", x="Predicted Plant Nitrogen (micromol/g)", y="Soil Temperature (°C)", color="Season")+
  theme_minimal()
```

**Discussion About Vegetation-Related Plant Nitrogen Plots**

From the plots, we see that the correlation of plant nitrogen and vegetation % or sward height is actually negative. The only correlation that matches our prediction is the negative correlation of plant nitrogen and soil temperature. The different seasons in each plot occupy a different area, with plants in spring usually having a larger vegetation area, taller plant height, and greater soil temperature at a significantly lower plant nitrogen than plants in autumn.

**Leaf and Plant Size as a Possible Indicator of Plant Nitrogen**

One of the important uses of nitrogen for plants is the creation of chlorophyll, a green pigment required for photosynthesis. Having a larger leaf area could mean a higher total amount of chlorophyll and a higher nitrogen content in a plant. A larger plant would also have a larger biomass and would require more nitrogen from the soil. We can test whether leaf and plant size are positive associated with plant nitrogen content.

```{r}
plant_joined %>% 
  select(predicted_n, total_leaf_area, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=total_leaf_area))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. Average Leaf Area", x="Predicted Plant Nitrogen (micromole/gram)", y="Average Leaf Area (mm^2)", color="Season")+
  theme_minimal()
```

```{r}
plant_joined %>% 
  select(predicted_n, plant_area, season) %>% 
  na.omit() %>% 
  ggplot(aes(x=predicted_n, y=plant_area))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)+
  labs(title="Predicted Plant Nitrogen vs. Plant Area", x="Predicted Plant Nitrogen (micromole/gram)", y="Plant Area (cm^2)", color="Season")+
  theme_minimal()
```


**Discussion About Leaf 


## Egg Amount Predictor Shiny App

**Testing QQ Residuals Plot**
```{r}
egg_calculator <- lm(formula=n_eggs ~ predicted_n + bare_ground + veg + sward_height + plant_area + average_leaf_size + temperature_ambient + temperature_soil, data=plant_joined)
egg_calculator_residuals <- egg_calculator$residuals
hist(egg_calculator_residuals)
qqnorm(egg_calculator_residuals)
qqline(egg_calculator_residuals)
```

**Print Linear Regression Model Summary**
```{r}
summary(egg_calculator)
```

```{r}

ui <- page_sidebar(
  title = "Egg Calculator Regression",
  sidebar = sidebar(
    
    # Select predictor variables
    checkboxGroupInput(
      "predictor_vars",
      "Select Predictor Variables:",
      choices = c("predicted_n",
                  "bare_ground",
                  "veg",
                  "sward_height",
                  "plant_area",
                  "average_leaf_size",
                  "temperature_ambient",
                  "temperature_soil"
      ),
      selected = c("predicted_n")
    ),
    
    # Action button to run regression
    actionButton("run_regression", "Run Regression", class = "btn-primary mt-3")
  ),
  
  # Main panel with plots and results
  navset_card_tab(
    nav_panel(
      "Scatter Plots",
      card(
        card_header("Egg Count vs Predictor Variables"),
        plotOutput("scatter_plots", height = "600px")
      )
    ),
    nav_panel(
      "Regression Summary",
      card(
        card_header("Model Summary"),
        verbatimTextOutput("regression_summary")
      ),
      card(
        card_header("Model Coefficients"),
        tableOutput("coefficients_table")
      )
    ),
    nav_panel(
      "Diagnostic Plots",
      card(
        card_header("Residual Diagnostics"),
        plotOutput("diagnostic_plots", height = "600px")
      )
    )
  )
)

server <- function(input, output, session) {
  
  # Fit regression model when button is clicked
  model <- eventReactive(input$run_regression, {
    req(input$predictor_vars)
    
    # Create formula with egg count as dependent variable
    formula_str <- paste("n_eggs ~", paste(input$predictor_vars, collapse = " + "))
    formula_obj <- as.formula(formula_str)
    
    # Fit linear model
    lm(formula_obj, data = plant_joined)
  })
  
  # Scatter plots
  output$scatter_plots <- renderPlot({
    req(input$predictor_vars)
    
    n_predictors <- length(input$predictor_vars)
    
    # Calculate grid dimensions
    n_cols <- ceiling(sqrt(n_predictors))
    n_rows <- ceiling(n_predictors / n_cols)
    
    par(mfrow = c(n_rows, n_cols), mar = c(4, 4, 2, 1))
    
    for (predictor in input$predictor_vars) {
      plot(plant_joined[[predictor]], plant_joined$n_eggs,
           xlab = predictor,
           ylab = "Egg Count",
           main = paste("Egg Count vs", predictor),
           pch = 19,
           col = rgb(0.2, 0.4, 0.6, 0.6),
           cex = 1.2)
      
      # Add regression line for visualization
      abline(lm(plant_joined$n_eggs ~ plant_joined[[predictor]]), 
             col = "red", lwd = 2)
    }
  })
  
  # Regression summary
  output$regression_summary <- renderPrint({
    req(model())
    summary(model())
  })
  
  # Coefficients table
  output$coefficients_table <- renderTable({
    req(model())
    
    coef_summary <- summary(model())$coefficients
    coef_df <- as.data.frame(coef_summary)
    coef_df$Variable <- rownames(coef_df)
    coef_df <- coef_df[, c("Variable", "Estimate", "Std. Error", "t value", "Pr(>|t|)")]
    rownames(coef_df) <- NULL
    
    coef_df
  }, digits = 4)
  
  # Diagnostic plots
  output$diagnostic_plots <- renderPlot({
    req(model())
    
    par(mfrow = c(2, 2))
    plot(model())
  })
}

shinyApp(ui, server)
```


## Limitations Of Data and Correlations
The data collected and used for the correlations in our study is specifically from the Shotover Country Park in Oxfordshire, United Kingdom. The only the plant species used in this data was Rumex acetosa or sorrel, which limits the study to butterflies that specialize in consuming this plant. This may impact the validity of our study because the data sample isn't representative of plants and butterflies throughout the world. Future studies on the trade-off between plant nutritional value and egg-laying preference could be done in other areas of the world, on other species of plants, and other species of butterflies.