---
title: "work_stanley"
output: html_document
---

## Libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(ggplot2)
library(naniar)
library(ggmap)
library(UKFE)
```

## Introduction and Motivation
In this project, we will study the factors that may affect plant nutritional quality and butterfly preference for laying eggs. Butterflies contribute greatly to the wide diversity of flower species by cross-pollination when it feeds on nectar. By learning what factors butterflies have a greater tendency to lay its eggs at, we could possibly predict butterfly migration patterns?(thinking about this) as well as advocate for butterfly habitat protection at crucial areas.

## Data
We will use data from [William B.V. Langdon, Richard Fox, Owen T. Lewis, Host plant use is driven by microclimate not nutritional quality in a grassland butterfly](https://doi.org/10.1101/2025.11.15.688506). The original research article and data sets are included in the data folder in this repository.
```{r}
egg_laying <- read_csv("data/egg_laying_data.csv") %>% 
  clean_names()
plant_chemistry <- read.csv("data/plant_chemistry_data.csv") %>%
  clean_names()
```


## Exploratory Data Analysis

**Looking Into Egg Laying Data Set**
```{r}
summary(egg_laying)
```

```{r}
egg_laying %>% 
  miss_var_summary() #check for NA's
```

```{r}
egg_laying %>% 
  gg_miss_var()+ #visualizing the NA's
  labs(title="NA's in Egg Laying Data Set")
```

Through the NA summary and plot, we can see the majority of NA's come from environment_date with most of the other NA's from leaf2 and leaf3 measurements. It's possible that they left the environmental measurement dates out on accident as variables besides ambient and soil temperature were measured despite the date being NA. For the leaf measurements, .


Since the egg_laying and plant_chemistry data sets have the same plant id, we can join the two datasets together using a join function. It's important to know that plant_chemistry also shares the same leaf length, width, area, plant area, and season, so we can select the unique variables of plant_chemistry.

```{r}
plant_joined <- left_join(egg_laying, plant_chemistry %>% select(plant_id, n, c, cn_percent, n_molar, c_molar, cn_molar, quality_pc1, quality_pc2), by="plant_id")
```

**Distribution of Important Variables**

**Creating Map for Data Set and Issues**

In addition to the packages we've included from lecture, we also added the package `UKFE` to convert the British national grid reference coordinates into latitude and longitude.

```{r}
coordtomap <- egg_laying %>% 
  select(gps) %>% 
  na.omit() %>% 
  separate(gps, into=c("SP", "grid"), sep=2) %>% # the SP in 
  separate(grid, into=c("easting", "northing"), sep=5) %>% 
  mutate(easting=as.numeric(easting)+400000) %>% 
  mutate(northing=as.numeric(northing)+200000) %>% 
  rowwise() %>% 
  mutate(coords=list(ConvertGridRef(c(easting,northing)))) %>% 
  unnest(coords)
```

## Factors that may influence nutritional quality

Butterfly offspring suitability can be determined primarily by two factors: environmental temperature and plant nutritional quality. 




```{r}
plant_joined %>% 
  select(temperature_ambient, predicted_n, season.x) %>% 
  na.omit() %>% 
  ggplot(aes(x=season.x, y=predicted_n))+
  geom_boxplot(mapping=aes(fill=season.x))
```

```{r}
 plant_joined %>% 
  ggplot(aes(x=temperature_ambient, color=season.x, fill=season.x))+
  geom_density()
```

```{r}
 plant_joined %>% 
  ggplot(aes(x=temperature_soil, color=season.x, fill=season.x))+
  geom_density()
```

```{r}
plant_joined %>% 
  select(predicted_n, n_molar) %>% 
  ggplot(aes(predicted_n, n_molar))+
  geom_point()
```

```{r}
plant_joined %>% 
  select(occupied, temperature_ambient, season.x) %>% 
  mutate(occupied=factor(occupied)) %>% 
  ggplot(aes(x=occupied))+
  geom_bar(mapping=aes(fill=season.x), color="black", position="dodge")
```

```{r}
plant_joined %>% 
  select(occupied, predicted_n, season.x) %>% 
  mutate(occupied=factor(occupied)) %>% 
  ggplot(aes(x=predicted_n))+
  geom_density(mapping=aes(fill=occupied), position="dodge", alpha=0.5)+
  facet_wrap(~season.x, ncol=2, labeller=label_both)
```

# Do butterflies prefer plants with higher percentage of bare land?

```{r}
egg_laying %>% 
  filter(season=="autumn") %>% 
  na.omit() %>% 
  ggplot(mapping=aes(x=bare_ground, y=temperature_soil))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)
```

```{r}
egg_laying %>% 
  filter(season=="spring") %>% 
  na.omit() %>% 
  ggplot(mapping=aes(x=bare_ground, y=predicted_n))+
  geom_point(mapping=aes(color=season))+
  geom_smooth(method="lm", se=T)
```

```{r}
plant_joined %>% 
  na.omit() %>% 
  mutate(occupied=factor(occupied)) %>% 
  ggplot(mapping=aes(x=bare))+
  geom_bar(mapping=aes(fill=occupied), position="dodge")
```

```{r}
coordtomap <- egg_laying %>% 
  select(gps) %>% 
  na.omit() %>% 
  separate(gps, into=c("SP", "grid"), sep=2) %>% 
  separate(grid, into=c("easting", "northing"), sep=5) %>% 
  mutate(easting=as.numeric(easting)+400000) %>% 
  mutate(northing=as.numeric(northing)+200000) %>% 
  rowwise() %>% 
  mutate(coords=list(ConvertGridRef(c(easting,northing)))) %>% 
  unnest(coords)
```

```{r}
coordtomap %>% 
  summary()
```

```{r}
#lat <- c(51.74 , 51.75)
#long <- c(-1.185 , -1.181)
#bbox <- make_bbox(long, lat, f=1000)
#map1 <- get_stadiamap(bbox, maptype="stamen_terrain", zoom=4)
#ggmap(map1)+
  #geom_point(data=coordtomap, aes(longitude, latitude), size=0.75, color="black", alpha=0.8)
```

```{r}
#egg_laying %>% 
  #mutate(larvae_egg_ratio=)
```


